A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-10 doc/popl2013/langdef.tex
additional changes

@@

$T$ & ::= &  & types & $bv$ & ::= & false $\;\;\;\mid\;\;\;$ true & \\
  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types & & & & \\
  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple & $iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ & \\
  &$\mid$& $T@(\oton{r})$ & pointer & & & & \\
  &$\mid$& $\text{coloring}(r)$ & region coloring & $e$ & ::= & & expressions \\
  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship &   &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions &   &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
& & & &   &$\mid$& $id$ &  \\
$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints &   &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
$\omega$ & ::= & $r_1 \leq r_2$ & subregion   &   
  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ & \\
  &$\mid$& $r_1 * r_2$ & disjointness &   
  &$\mid$ & $\text{downregion}(e, r_1,\ldots,r_n)$ & \\  
& & &  & 
  &$\mid$& $\text{read}(e_1)$ $\;\mid\;$ $\text{write}(e_1, e_2)$ & memory access \\
$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges & 
  & $\mid$ & $\text{reduce}(id, e_1, e_2)$ & \\
$\phi$ & ::= & reads$(r)$ $\;\mid\;$ writes$(r)$ $\;\mid\;$ $\text{reduces}_{id}(r)$ &  &  
  &$\mid$& $\text{newcolor}\ r$ $\;\mid\;$ $\text{color}(e_1, e_2, e_3)$ & coloring \\
& & & &  
  &   $\mid$& $e_1 + e_2$ & integer ops \\
$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes  &   
       & $\mid$& $e_1 < e_2$ & comparisons \\
$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & &   
  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
& & & &   
       & $\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
$v$ & ::= & & values &   
       & $\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values &   
       & $\mid$ & $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple & 
       & $\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address & 
       & $\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring & 
    & & & \\
  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & reg. relation instance & 
   & & & \\
%$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
%\\
%$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
%\\
%$e$ & ::= & & expressions \\
%  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
%  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
%  &$\mid$& $id$ &  \\
%  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
%  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
%  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
%  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
%  &$\mid$& $e_1 + e_2$ & integer operations \\
%  &$\mid$& $e_1 < e_2$ & comparison operations \\
%  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
%  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
%  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
%  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
%  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
%  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
\end{tabular}
}
\caption{Core Legion}
\label{fig:langdef}
\end{figure*}

@@

Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
that still illustrates all of the important issues.  Types
include booleans, integers, tuples, and pointers.  Note that pointers
are annotated with a list of regions into which they may point; in a
well-typed program, the pointer is guaranteed to point to at least one
of those regions. There is a special type for {\em colorings}, which
are used to specify partitions of regions into subregions.
Functions/tasks in Legion are named, accept one or more arguments and
return a value of some type.  A function also declares the region
access privileges and coherence it requires.  Function types are
universally quantified over all region names appearing in the type.
The final type is a region relationship, instance of which capture a value and
one or more regions satisfying associated constraints.  An
instance of a region relation can be read out of the heap and
unpacked, giving new local names to the regions contained in
the instance.

@@

\vdash \{ & \text{task }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, \\

@@

          & \text{task }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet

@@

\infax[BOOL]{\typeenv{bv}{bool}} & 
\infax[INT]{\typeenv{iv}{int}} & 
\infrule[TUPLE1]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1} & 
\infrule[TUPLE2]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2} 

@@

\infrule[VAR]{\Gamma(id) = T}{\typeenvx id : T} &
\infax[NULL]{\typeenvx \text{null }T@r : T@r} &
\infrule[ISNULL]{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool} &
\infrule[MAKETUPLE]{\typeenvx e_1 : T_1 \andalso \typeenvx e_2 : T_2}{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}

@@

\infax[NEW]{\typeenvx \text{new }T@r : T@r} &
\infax[NEWCOLOR]{\typeenvx \text{newcolor }r : \text{coloring}(r)} &
\infrule[COLORING]{\typeenvx e_1 : \text{coloring}(r) \andalso \typeenvx e_2 : T@r \andalso \typeenvx e_3 : int}{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}

@@

\infrule[LET]{\typeenvx e_1 : T_1 \\ \typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2}{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} &
\infrule[ADD]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 + e_2 : int} &
\infrule[COMPARE]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 < e_2 : bool}

@@

\infrule[UPREG]{\typeenvx e : T@(r'_1, \ldots r'_k) \andalso \forall i. \exists j, r'_i \leq r_j \in \Omega^*}{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})} &
\infrule[DOWNREG]{\typeenvx e : T@(r'_1, \ldots r'_k)

@@

\infrule[READ]{\typeenvx e_1 : T@(\oton{r}) \\ \forall i, reads(r_i) \in \Phi^*}{\typeenvx \text{read}(e_1) : T} &
\infrule[WRITE]{\typeenvx e_1 : T@(\oton{r}) \\ \typeenvx e_2 : T \\ \forall i, writes(r_i) \in \Phi^*}{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})} &
\infrule[REDUCE]{\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \andalso \typeenvx e_2 : T_2 \\

@@

\infrule[PACK]{T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\ \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\ \typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]}{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} &
\infrule[IF-ELSE]{\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} &
\infrule[TASK-CALL]{\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\ \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\ \Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*}{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}

@@

\infrule[UNPACK]{T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \andalso \typeenvx e_1 : T_1 \\

@@

\infrule[PARTITION]{\typeenvx e_1 : \text{coloring}(r_p) \\ \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\ \typeenvx[O=\Omega'] e_2 : T \\ \{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset}

@@

\infrule[PROGRAM]{\text{for }1 \leq i \leq p, \\

@@

  {\vdash \{  \text{task }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, 

@@

           \text{task }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet}

@@

\item $H$ is a {\em heap typing}, assigning an invariant type to every memory location.

@@

\item $mark\_coherence(E, \hat Q)$ records the coherence of operations in trace $E$ according to coherence annotations $Q$.

\item $valid\_interleave(S, C, E', \oton{E})$ is true if $E'$ is a valid interleaving
of the memory traces $\oton{E}$.
\end{itemize}
Figure~\ref{fig:opsemfns} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
{\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
is at least an interleaving, containing each constituent trace as a subequence and no other operations.

Some rules modeling concurrent execution are locally
non-deterministic, guessing values that result from
concurrent tasks.  For example, the rule for {\tt read(e)} returns an
arbitrary value $v'$ of the correct type if the location being read is
in the clobber set $C$, the set of locations concurrent tasks may
clobber (i.e., update).\footnote{Note that this is also the reason for
including a global heap typing.}  The guess is eventually checked for
consistency by the $valid\_interleave$ predicate of some enclosing
expression, at a point where the task that wrote $v'$ is also local to
the expression being evaluated.

@@

The most involved rule in Core Legion's semantics is for task calls.
Note that the only regions accessible to the callee (through $M'$) are the region arguments.
Also, if any regions are used with relaxed coherence (either {\em atomic} or {\em simult} instead of
{\em excl}), those locations are added to the clobber set, allowing the
impact of concurrent modifications of the store to be handled by the caller.  
%This is done by
%annotating the called function's memory trace with the relaxed coherence on a per-location basis
%and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.

\begin{figure*}
\centering{
$
\begin{array}{lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \text{ where } S' = apply(S, E) \\
mark\_coherence([], \hat Q) & = & [] \\
mark\_coherence(E \dplus [ op(l, c, v) ], \hat Q) & = & mark\_coherence(E, \hat Q) \dplus \\
& & op(l, c', v), \text{ where } \\
& & c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases} \\
any\_interleave([], [], \ldots, []) & = & true \\
any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfns}
\end{figure*}

@@

\text{task }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\

@@

\infax[OP-BOOL]{\opsenvx bv \mapsto bv, []} &
\infax[OP-INT]{\opsenvx iv \mapsto iv, []} &
\infax[OP-NULL]{\opsenvx \text{null }T@r \mapsto null, []} 

@@

\infrule[OP-MAKETUPLE]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S' = apply(S, E_1) \\

@@

\infrule[OP-TUPLE1]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.1} \mapsto v_1, E} &
\infrule[OP-TUPLE2]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.2} \mapsto v_2, E}

@@

\infrule[OP-NEW]{l \in M(r) \\ H(l) = M \llbracket T \rrbracket}{\opsenvx \text{new }T@r \mapsto l, []} &
\infrule[OP-VAL]{L(id) = v}{\opsenvx id \mapsto v, []} &
\infrule[OP-ISNULL-FALSE]{\opsenvx e \mapsto l, E}{\opsenvx \text{isnull}(e) \mapsto \textit{false}, E} &
\infrule[OP-ISNULL-TRUE]{\opsenvx e \mapsto null, E}{\opsenvx \text{isnull}(e) \mapsto true, E}

>>>>

% override BCP's typesetrule function to left-justify multiple lines of hypotheses
\renewcommand{\typesetrule}[2]{%
   \setrulebody{%
      \frac{\begin{array}{@{}l@{}}#1\end{array}}%
           {\begin{array}{@{}l@{}}#2\end{array}}}}

@@

$T$ & ::= &  & types & $bv$ & ::= & false $\;\;\;\mid\;\;\;$ true & \\
  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types & & & & \\
  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple & $iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ & \\
  &$\mid$& $T@(\oton{r})$ & pointer & & & & \\
  &$\mid$& $\text{coloring}(r)$ & region coloring & $e$ & ::= & & expressions \\
  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship &   &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions &   &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
& & & &   &$\mid$& $id$ &  \\
$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints &   &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
$\omega$ & ::= & $r_1 \leq r_2$ & subregion   &   
  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ & \\
  &$\mid$& $r_1 * r_2$ & disjointness &   
  &$\mid$ & $\text{downregion}(e, r_1,\ldots,r_n)$ & \\  
& & &  & 
  &$\mid$& $\text{read}(e_1)$ $\;\mid\;$ $\text{write}(e_1, e_2)$ & memory access \\
$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges & 
  & $\mid$ & $\text{reduce}(id, e_1, e_2)$ & \\
$\phi$ & ::= & reads$(r)$ $\;\mid\;$ writes$(r)$ $\;\mid\;$ $\text{reduces}_{id}(r)$ &  &  
  &$\mid$& $\text{newcolor}\ r$ $\;\mid\;$ $\text{color}(e_1, e_2, e_3)$ & coloring \\
& & & &  
  &   $\mid$& $e_1 + e_2$ & integer ops \\
$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes  &   
       & $\mid$& $e_1 < e_2$ & comparisons \\
$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & &   
  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
& & & &   
       & $\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
$v$ & ::= & & values &   
       & $\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & function calls \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values &   
       & $\mid$ & $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple & 
       & $\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address & 
       & $\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring & 
    & & & \\
  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & reg. relation instance & 
   & & & \\
%$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
%\\
%$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
%\\
%$e$ & ::= & & expressions \\
%  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
%  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
%  &$\mid$& $id$ &  \\
%  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
%  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
%  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
%  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
%  &$\mid$& $e_1 + e_2$ & integer operations \\
%  &$\mid$& $e_1 < e_2$ & comparison operations \\
%  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
%  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
%  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & function calls \\
%  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
%  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
%  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
\end{tabular}
}
\caption{Core Legion}
\label{fig:langdef}
\end{figure*}

@@

Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
that still illustrates all of the important issues.  Types
include booleans, integers, tuples, and pointers.  Note that pointers
are annotated with a list of regions into which they may point; in a
well-typed program, the pointer is guaranteed to point to at least one
of those regions. There is a special type for {\em colorings}, which
are used to specify partitions of regions into subregions.
Functions in Legion are named, accept one or more arguments and
return a value of some type.  A function also declares the region
access privileges and coherence it requires.  A function may be executed
within the current task or run as a separate subtask - this decision is made
by the Legion task scheduler.  Function types are
universally quantified over all region names appearing in the type.
The final type is a region relationship, an instance of which captures a value and
one or more regions satisfying associated constraints.  A region relation instance
can be written into the heap and later read and
unpacked, giving new local names to the regions contained in
the instance.

@@

\vdash \{ & \text{function }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, \\

@@

          & \text{function }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet

@@

\infax[T-Bool]{\typeenv{bv}{bool}} & 
\infax[T-Int]{\typeenv{iv}{int}} & 
\infrule[T-Tuple1]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1} & 
\infrule[T-Tuple2]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2} 

@@

\infrule[T-Var]{\Gamma(id) = T}{\typeenvx id : T} &
\infax[T-Null]{\typeenvx \text{null }T@r : T@r} &
\infrule[T-IsNull]{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool} &
\infrule[T-MakeTuple]{\typeenvx e_1 : T_1 \andalso \typeenvx e_2 : T_2}{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}

@@

\infax[T-New]{\typeenvx \text{new }T@r : T@r} &
\infax[T-Newcolor]{\typeenvx \text{newcolor }r : \text{coloring}(r)} &
\infrule[T-Color]{\typeenvx e_1 : \text{coloring}(r) \andalso \typeenvx e_2 : T@r \andalso \typeenvx e_3 : int}{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}

@@

\infrule[T-Let]{\typeenvx e_1 : T_1 \\ \typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2}{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} &
\infrule[T-Add]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 + e_2 : int} &
\infrule[T-Compare]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 < e_2 : bool}

@@

\infrule[T-UpRgn]{\typeenvx e : T@(r'_1, \ldots r'_k) \andalso \forall i. \exists j, r'_i \leq r_j \in \Omega^*}{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})} &
\infrule[T-DownRgn]{\typeenvx e : T@(r'_1, \ldots r'_k)

@@

\infrule[T-Read]{\typeenvx e_1 : T@(\oton{r}) \\ \forall i, reads(r_i) \in \Phi^*}{\typeenvx \text{read}(e_1) : T} &
\infrule[T-Write]{\typeenvx e_1 : T@(\oton{r}) \\ \typeenvx e_2 : T \\ \forall i, writes(r_i) \in \Phi^*}{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})} &
\infrule[T-Reduce]{\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \andalso \typeenvx e_2 : T_2 \\

@@

\infrule[T-Pack]{T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\ \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\ \typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]}{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} &
\infrule[T-IfElse]{\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} &
\infrule[T-Call]{\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\ \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\ \Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*}{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}

@@

\infrule[T-Unpack]{T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \andalso \typeenvx e_1 : T_1 \\

@@

\infrule[T-Part]{\typeenvx e_1 : \text{coloring}(r_p) \\ \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\ \typeenvx[O=\Omega'] e_2 : T \\ \{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset}

@@

\infrule[T-Program]{\text{for }1 \leq i \leq p, \\

@@

  {\vdash \{  \text{function }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, 

@@

           \text{function }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet}

@@

\item $H$ is a {\em heap typing}, assigning an invariant type to every memory location.  The heap typing
serves two important roles.  First, it is used to bound the values that can be used for reads when 
concurrent operation may make the exact value uncertain.  Second, it greatly simplifies the consistency
check of pointer values, allowing a simple test against the heap typing for a location in lieu of a
potentially unbounded walk through the store for complicated pointer data structures.

@@

\item $mark\_coh(E, \hat Q)$ records the coherence of operations in trace $E$ according to coherence annotations $\hat Q$.

\item $valid\_interleave(S, C, E', \oton{E})$ is true if $E'$ is a valid interleaving
of the memory traces $\oton{E}$.
\end{itemize}
Figure~\ref{fig:opsemfns} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
{\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
is at least an interleaving, containing each constituent trace as a subsequence and no other operations.

Some rules modeling concurrent execution are locally
non-deterministic, guessing values that result from
concurrent tasks.  For example, the rule for {\tt read(e)} returns an
arbitrary value $v'$ of the correct type if the location being read is
in the clobber set $C$, the set of locations concurrent tasks may
clobber (i.e., update).  The guess is eventually checked for
consistency by the $valid\_interleave$ predicate of some enclosing
expression, at a point where the task that wrote $v'$ is also local to
the expression being evaluated.

@@

The most involved rule in Core Legion's semantics is for function calls.
Note that the only regions accessible to the callee (through $M'$) are the region arguments.
Also, if any regions are used with relaxed coherence (either {\em atomic} or {\em simult} instead of
{\em excl}), those locations are added to the clobber set, allowing the
impact of concurrent modifications of the store to be handled by the caller.  
%This is done by
%annotating the called function's memory trace with the relaxed coherence on a per-location basis
%and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.

\begin{figure*}
\centering{
$
\begin{array}{@{}l@{}}
\begin{array}{@{}l@{\hspace{-0.25in}}l@{}} 
\begin{array}{@{}lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
& & \text{ where } S' = apply(S, E)
\end{array} &
\begin{array}{@{}ll}
mark\_coh([], \hat Q) & = [] \\
mark\_coh([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coh(E, \hat Q), \\
\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases}}
\end{array}
\end{array} \\
\begin{array}{@{}lll}
any\_interleave([], [], \ldots, []) & = & true \\
any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
\end{array}
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfns}
\end{figure*}

@@

\text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\

@@

\infax[E-Bool]{\opsenvx bv \mapsto bv, []} &
\infax[E-Int]{\opsenvx iv \mapsto iv, []} &
\infax[E-Null]{\opsenvx \text{null }T@r \mapsto null, []} 

@@

\infrule[E-MakeTuple]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S' = apply(S, E_1) \\

@@

\infrule[E-Tuple1]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.1} \mapsto v_1, E} &
\infrule[E-Tuple2]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.2} \mapsto v_2, E}

@@

\infrule[E-New]{l \in M(r) \\ H(l) = M \llbracket T \rrbracket}{\opsenvx \text{new }T@r \mapsto l, []} &
\infrule[E-Var]{L(id) = v}{\opsenvx id \mapsto v, []} &
\infrule[E-IsNull-F]{\opsenvx e \mapsto l, E}{\opsenvx \text{isnull}(e) \mapsto \textit{false}, E} &
\infrule[E-IsNull-T]{\opsenvx e \mapsto null, E}{\opsenvx \text{isnull}(e) \mapsto true, E}
